Generating Correlated Data for Omics Simulation

### 2024-09-24 manuscript/paper.qmd
Add some references

<!-- TODO which dataset do we want to use here? -->
We used mouse liver RNA-seq data set from accession GSE77221[@Weger2019] where we expected a significant amount of gene-gene correlation due to the samples having been collected throughout the course of a day.
We used this data to prime our simulation and compared real and simulated data sets, both with 12 samples, to each other.
We repeated the simulations a total of 8 times to estimate variance.
The simulations were then repeated without any dependence for comparison.

@@

We benchmark DESeq2 <!-- TODO cite -->, a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both. We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes. We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 20 "experimental" samples each with and and without dependence.

@@

We further use our simulation method to benchmark CYCLOPS <!-- TODO cite -->, which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures. We use a mouse liver time series data set (GSE151565, Aaronson et al. 2020), which contains a total of 77 samples every 3 hours, for 36 hours. We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point. We then use these to simulate 20 time series data sets with rank $2$ dependence structure, each with 8 time points and 6 samples per time point, and 20 such data sets without dependence.

@@

Specifically, we do:
1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c, \ldots, c$.
4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
3. Set $D = cI$

@@

The `corpcor` package <!-- TODO cite --> computes a James-Stein type shrinkage estimator for the covariance matrix.
For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
It computes optimal values of $\lambda_1$, and $\lambda_2$, its two regularization coefficients.
It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D + UWU^T$.

This algorithm is:
1. Compute the $\lambda_1$ and $\lambda_2$ values from `corpcor::estimate.lambda` and `corpcor::estimate.lambda.var` functions on $Z$, respectively.
2. Set $D$ to be diagonal with $D_{ii} = \sqrt{\lambda_1} (\lambda_2 \sigma_{med} + (1 - \lambda_2) \sigma_i)$ where $\sigma_i$ is the standard deviation of the $Z_{i\cdot}$ and $\sigma_{med}$ is the median of the $\sigma_i$.
3. Set $U$ to be $\sqrt{1-\lambda} S Z / \sqrt{n-1}$ where $S$ is the diagonal matrix with $S_{ii} = \lambda_2 \sigma_{med} / \sigma_i + (1 - \lambda_2)$.
4. Set $W$ to the identity.

@@

The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) <!-- TODO cite --> provides a dedicated RNA-seq and scRNA-seq simulator using a NORTA approach to simulate gene dependence.
In contrast to this package, it uses WGCNA to determine the correlation matrix, which is a gene network approach.
However, this method takes significant computational resources.
Indeed, the `SPsimSeq` paper generated data for just 5000 genes based on a randomly sampled 5000 gene subset of the RNA-seq data and our attempts to use `SPsimSeq` to generate a full sample exhausted the memory of a 24GB computer.
In contrast, our method runs in seconds to generate a 40,000 gene samples on the same computer.
`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
In contrast, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator <!-- TODO cite --> for scRNA-seq uses NORTA and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
However, it limits this correlation matrix to top-expressed genes.

>>>>

<!-- TODO which dataset do we want to use here? -->
We used mouse liver RNA-seq data set from accession GSE77221 [@Weger2019] where we expected a significant amount of gene-gene correlation due to the samples having been collected throughout the course of a day.
We used this data to prime our simulation and compared real and simulated data sets, both with 12 samples, to each other.
We repeated the simulations a total of 8 times to estimate variance.
The simulations were then repeated without any dependence for comparison.

@@

We benchmark DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both. We use a fly whole body RNA-Seq data set GSE81142 and select samples of male flies without treatment and after at least 2 hours of feeding to simulate 20 control samples with rank $k=2$ dependence structure, and 20 control samples assuming independence of the genes. We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 20 "experimental" samples each with and and without dependence.

@@

We further use our simulation method to benchmark CYCLOPS [@Anafi2017], which infer relative times for a set of unlabeled samples using an autoencoder to identify circular structures. We use a mouse liver time series data set (GSE151565, Aaronson et al. 2020), which contains a total of 77 samples every 3 hours, for 36 hours. We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point. We then use these to simulate 20 time series data sets with rank $2$ dependence structure, each with 8 time points and 6 samples per time point, and 20 such data sets without dependence.

@@

Specifically, we do:

1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c, \ldots, c$.
4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
3. Set $D = cI$

@@

The `corpcor` package [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] computes a James-Stein type shrinkage estimator for the covariance matrix.
For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
It computes optimal values of $\lambda_1$, and $\lambda_2$, its two regularization coefficients.
It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D + UWU^T$.

This algorithm is:

1. Compute the $\lambda_1$ and $\lambda_2$ values from `corpcor::estimate.lambda` and `corpcor::estimate.lambda.var` functions on $Z$, respectively.
2. Set $D$ to be diagonal with $D_{ii} = \sqrt{\lambda_1} (\lambda_2 \sigma_{med} + (1 - \lambda_2) \sigma_i)$ where $\sigma_i$ is the standard deviation of the $Z_{i\cdot}$ and $\sigma_{med}$ is the median of the $\sigma_i$.
3. Set $U$ to be $\sqrt{1-\lambda} S Z / \sqrt{n-1}$ where $S$ is the diagonal matrix with $S_{ii} = \lambda_2 \sigma_{med} / \sigma_i + (1 - \lambda_2)$.
4. Set $W$ to the identity.

@@

The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and scRNA-seq simulator using a NORTA approach to simulate gene dependence.
In contrast to this package, it uses WGCNA to determine the correlation matrix, which is a gene network approach.
However, this method takes significant computational resources.
Indeed, the `SPsimSeq` paper generated data for just 5000 genes based on a randomly sampled 5000 gene subset of the RNA-seq data and our attempts to use `SPsimSeq` to generate a full sample exhausted the memory of a 24GB computer.
In contrast, our method runs in seconds to generate a 40,000 gene samples on the same computer.
`SPsimSeq` is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options.
In contrast, our `dependentsimr` package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

The [scDesign2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02367-2) simulator [@Sun2021] for scRNA-seq uses NORTA and, like our method, uses the approach of estimating the correlation matrix from the normalized dataset.
However, it limits this correlation matrix to top-expressed genes.
